### 2019-10-02 ThesisJo/2_background.tex
2019-10-02

Imagine an array with $M$ sensors with a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simply the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
\end{figure}
%
The basic idea behind MVDR is to compute the set of weights that minimizes noise and signal power accumulated by the array, while ensuring unit gain for signals from the pixel of interest. With pre-delayed data this can be formulated as follows:
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
This optimization problem has the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
\end{gather}
%
where $\dvec1$ is a row vector of ones representing broadside steering and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array. We estimate $\R$ by computing a sample covariance matrix $\eR$. A good estimate requires some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

>>>>

Imagine an array with $M$ sensors with a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we can simplify the notation to $x_m[\theta,n-\Delta_m] = x_m[n]$.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
\end{figure}
%
The basic idea behind MVDR is to compute the set of weights that minimizes noise and interference accumulated by the array, while ensuring unit gain for signals coming from the pixel of interest. With pre-delayed data this can be formulated as follows:
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
This optimization problem has the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
\end{gather}
%
where $\dvec1$ is a row vector of ones representing broadside steering and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array. We estimate $\R$ by computing a sample covariance matrix $\eR$. A good estimate requires some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

### 2019-10-02 ThesisJo/3_publications.tex
2019-09-30
2019-10-02

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Article relations. The first three articles reconstructs images from acoustic echoes. The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image. The two key traits are: accelerated adaptivity.  }
\end{figure}

@@

This article accelerates time-domain MVDR with GPUs for use in active sonar imaging. Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity while. We refer to it as fully adaptive because it computes array weights analytically from spatial features of the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

>>>>

\begin{figure}[tp]\label{III_fig_article_relation}
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Article relations (repost of Figure \ref{I_fig_article_relation} in chapter 1). The first three articles reconstructs images from acoustic echoes. The last article synthesizes image templates from 3D models and object parameters estimated from the reconstructed image. Accelerated adaptivity is key: it enables rapid dynamic responses to mission events. }
\end{figure}

@@

This article accelerates time-domain MVDR with GPUs for use in active sonar imaging. Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

### 2019-10-02 ThesisJo/4_discussion.tex
2019-10-02

\textbf{Paper I} - \textit{Time-domain MVDR beamforming on GPUs in active sonar imaging}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}

Time-domain MVDR is---in theory---the optimal way of computing weights that minimize variance (noise) in the image.

Time-domain MVDR is the reference method for creating sonar images using the minimum variance criterion. It computes array weights analytically, yielding optimal weights in the minimum variance sense. Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality. GPU acceleration mitigates the high computational complexity resulting from estimating and inverting a spatial covariance matrix, yielding real-time performance.

@@

>>>>

\textbf{Paper I} - \textit{Time-domain MVDR beamforming on GPUs in active sonar imaging}:
\begin{enumerate}
\item Accelerates the MVDR beamformer through algorithmic reductions and fine-tuning to a GPU, yielding a real-time solution.
\item Demonstrates speeds exceeding 1\,Mpixel/s on systems with no more than 32 channels, two orders of magnitude faster than a semi-optimized C-implementation on a CPU.
\item Offers rarely documented insights into hardware limitations, design implications and computational efficiency.
\end{enumerate}
% 
% Time-domain MVDR is---in theory---the optimal way of computing weights that minimize variance (noise) in the image. 
%

@@

MVDR is the reference method for creating sonar images using the minimum variance criterion. It computes array weights analytically, yielding optimal weights in the minimum variance sense. Initially, this may seem ideal, but its use is cumbersome: parameters for temporal averaging, spatial averaging and regularization must often be fine-tuned ad-hoc to obtain improved image quality. When using time-domain MVDR in systems with less than 32 channels, GPU acceleration mitigates the high computational complexity in estimating and inverting a spatial covariance matrix. For systems with narrow-band, narrow-beam, or both, the fre , yielding real-time performance.

reconstruction/Beamforming:
%
\begin{itemize}
\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
%
\item \emph{Solution space} is continuous for MVDR and discrete for LCA, but similarly shaped. 
%
\item \emph{Model complexity} has a sweet spot, a balance point where its estimates offer an ideal compromise between bias and variance. MVDR depend on 
%
\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both.
\end{itemize}

synthesis/simulation:

%A classification model should be neither too simple (high bias or underfitted)
%nor complex (high variance or underfit), but strike a balance between the
%two that captures statistically significant details of the object but not the
%noise. Higher detail-to-noise ratios favor more complex models and occur at

### 2019-10-02 ThesisJo/abstract.tex
2019-10-02

\input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\chapter{Abstract}

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\ifMonolithic\else\input{subfooter}\fi

>>>>

% \input{variables}\ifMonolithic\else\input{header}\input{subheader}\fi
%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

The world is highly dynamic. Any device aspiring to master it must be situationally perceptive and reactive. This requires numerous sensors gathering high quality data at high throughput, adaptive signal processing algorithms and powerful parallel hardware capable of delivering swift responses. In our case of sonar-equipped autonomous underwater vehicles and clinical ultrasound devices, this process revolves around acoustic images.

This thesis explores methods to accelerate adaptive acoustic image reconstruction and synthesis, through beamforming and simulation, respectively. Beamforming is the main attraction: the art of delaying and weighting sensor arrays to craft desired spatial responses. When aiming to minimize image noise and interference, two good candidates are the minimum variance distortionless response (MVDR) and low complexity adaptive (LCA) beamformer. We study time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated guessing), and conclude that:
%
\begin{itemize}
\item \emph{Image quality} is similar for our adaptive beamformers, but improved upon compared to static methods.
%
\item \emph{Solution space} is continuous for MVDR and discrete for LCA, but similarly shaped. 
%
\item \emph{Model complexity} has a sweet spot, a balance point where its estimates offer an ideal compromise between bias and variance. MVDR depend on 
%
\item \emph{Acceleration} of MVDR through algorithmic reductions and implementation on graphics processing units (GPUs) yields real-time responses in time-domain up to 32 channels and in frequency-domain for any system size with narrow-band, narrow-beam, or both. LCA is inherently fast and extremely well suited for acceleration, 
\end{itemize}
%
Finally, we perform image synthesis of common objects in the seabed with a GPU accelerated simulator. These are in turn used to perform adaptive automatic target recognition (ATR). The result is a more dynamic and flexible system yielding improved image quality, image quality and classification performance. Finally, we describe this system with a new navigational notation system, extended with topological concepts from robotics and homogeneous coordinates. It helps keeping track of complex geometries, numerical representations and higher level abstractions such as manipulating rigid objects.

%
%
%
%brings most adaptivity to the table, but also greatest computational load and model complexity.
%
%
%
%
%computational complexity increases but can be handled through algorithmic reduction and hardware acceleration, , for the adaptive beamformers are better than static methods, but ; \emph{user friendliness} is better for LCA than MVDR; 
%
%
%MVDR yields weights from a continuous solution space, but requires ad-hoc tuning of robustification parameters, is harder to understand and has high computational complexity. 
%
%The reconstruction process relies heavily on beamforming: the art of delaying and weighting sensor arrays to achieve desired spatial responses. Our adaptive beamformer reference, minimum variance distortionless response (MVDR), computes the weights analytically from an estimate of the full-array covariance matrix. 
%
%
%
%, both in time- and frequency domain. Then we ,  begin our study of a reference method  MVDR, frequency-domain MVDR and LCA all offer slight improvement in image quality over static methods. 
%
%All methods implemented on the GPU except LCA.
%
%A beamformer can be viewed as a spatial filter or model, and works best---as any model---when it is neither too simple nor complex. 
%
%
%
%We begin by accelerating the minimum variance distortionless response (MVDR) beamformer in time-domain, then  in both time- and frequency-domain, frequency-domain MVDR and low-complexity adaptive (LCA). 
% 
%
%All beamformers are fairly well suited for GPU acceleration because each pixel in the image can be processed independently.
%
% offer is considered the reference among adaptive beamformers, it offers neglible benefits over 
%
%sensory faculty
%
%as adaptive as possible with the least possible computational load.
%
%
%gather information
%dynamically analyze it
%and react to it
%
%adapt
%persist
%prevail
%
%Adaptive algorithm and the computation power to drive it.
%
%Cognition is the mental action or process of acquiring knowledge and understanding through thought, experience and the senses. 
%
%The intelligence and autonomy of acoustical imaging devices require more sensory data, algorithmic adaptivity and processing power. All are vital and equally important. 
%
%improves with more high quality sensory data, sophisticated adaptive signal processing and the hardware resources to deal with it all. 
%
%more sensory data, algorithmic adaptivity and processing power.
%
%autonomy improves with data amount, adaptive processing, processing power.
%
%information gathering and processing. 
%
%Key challenges that are addressed include active sonar and
%
%- Active sonar and clinical ultrasound
%- Different ways to achieve adaptivity. Analytical assessment of the data, computing weights. Approximating it. Trial-and-error. Different means to the same adaptivity
%- Computational complexity
%- Hardware acceleration
%- 
%
%Spatial sensitivity is like a balloon; squeezing it on one side inflates it on the other. 
%
%

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% \ifMonolithic\else\input{subfooter}\fi
